Solution of an Acoustic Transmission Inverse Problem by Extended Inversion

### 2020-12-15 repro/papers/discrep/discrep/paper.tex
tar dir prep for SEM to use overleaf
overleaf 20.12.14

\title{Solution of an Acoustic Transmission Inverse Problem by Extended Inversion}
\author{William W. Symes, Orcas Island, Huiyi Chen* and Susan E. Minkoff, The University of Texas at Dallas}

@@

\section{Introduction}
Model-based parameter estimation by least squares data fitting is a
widely used and successful technique for data analysis in science and
engineering \cite[]{Bertero:98,Vogel:02}, particularly in the
geosciences \cite[]{Parker:94,Tarantola:05}. In its application to
seismology, least-squares data fitting has come to be known as Full
Waveform Inversion (FWI), and is now well-established as a useful tool
for probing the earth's subsurface
\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. However, in its
application to seismology, FWI encounters a serious
practical obstacle, so-called ``cycle-skipping''. This term signifies
the tendency of the mean-square error function to exhibit many
stationary points far from the global minimizer (we
show an explicit example of this phenomenon later in this paper). Because of the
typical dimensions of field FWI problems, only local gradient-based
minimization algorithms are computationally practical. These find
stationary points (or local minima), so may stagnate at suboptimal and geologically
uninformative earth models. Local descent methods avoid suboptimal
stagnation only if initial models are already quite close to optimal,
in the sense of predicting the arrival times of seismic events to
within a small multiple of the dominant data wavelength
\cite[]{GauTarVir:86,VirieuxOperto:09,Plessix:10}.

@@

\begin{corollary}
  \label{thm:epjnonoise}
  Assume the hypotheses of Proposition \ref{thm:norminvexp}, item
  3. Then noting that $\|d\| = \|w_*\|/(4 \pi r)$
\begin{equation}
  \label{eqn:residnorm}
  e[m,\aw[m,d];d] 
= \frac{\alpha^4}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
\end{equation}
\begin{equation}
  \label{eqn:anninorm}
  p[m,\aw[m,d];d] = \frac{(4\pi r)^2}{2\|w_*\|^2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
    a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
\end{equation}
so 
\begin{equation}
\label{eqn:expjnonoise}
\tJa[m;d] = \frac{(4\pi r \alpha)^2}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
  a(t-(m-m_*)r)^2)^{-1}w_*(t)^2. 
\end{equation}
Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
\begin{equation}
  \label{eqn:dexpjnonoise}
  \frac{d}{dm}\tJa[m;d] = -\frac{r (4\pi r \alpha)^2}{\|w_*\|^2} \int \,dt \, 
  \left(a\frac{da}{dt}\right)(t-(m-m_*)r)(1+(4\pi r)^2 \alpha^2 
  a(t-(m-m_*)r)^2)^{-2}w_*(t)^2. 
\end{equation}
\end{corollary}

@@

\begin{proof}
  From the assumption $w_* \in W_{\mu}$ and Corollary
  \ref{thm:rampreallygood}, $|(m_{\infty}-m_*)r|\le \mu$. From the
  identity \ref{eqn:solnonoise},
  $\mbox{supp }\aw[m_{\infty};d] \subset
  [(m_{\infty}-m_*)r-\mu,(m_{\infty}-m_*)r+\mu] \subset
  [-2\mu,2\mu]$. Because of the support limitation, $a(t)=|t|$ in the
  interval of integration appearing in \ref{eqn:residnorm}, so
\[
  e[m_{\infty},\aw[m_{\infty},d];d] 
= 8 \pi^2 r^2 \alpha^4\int^{\mu}_{-\mu}\,dt\,|t-(m_{\infty}-m_*)r|^4(1+(4\pi r)^2 \alpha^2 
|t-(m_{\infty}-m^*)r|^2)^{-2}w_*(t)^2.
\]
and therefore
\begin{equation}
  \label{eqn:estresidnorm}
e[m_{\infty},\aw[m_{\infty};d];d] \le 8 \pi^2 r^2 (2\mu\alpha)^4 \|w_*\|^2 =
\frac{1}{2}(8 \pi r \mu \alpha)^4 \|d\|^2
\end{equation}
\end{proof}

@@

%%%%%%%%%%%
\begin{proposition}
  \label{thm:wnoiseres}
  Asumme that
  \begin{itemize}
  \item[1. ] $\alpha, \mu> 0$,
  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
  \item[3. ] $d_* = F[m_*]w_*$,
  \item[4. ] $n \in D$ and $d = d_* + n$.
  \end{itemize}
  Set $\eta = \|n\|/\|d_*\|$, and assume further that
  \begin{itemize}
  \item[5. ]
    \begin{equation}
      \label{eqn:mnoisecond}
      \eta < \frac{\sqrt{5}-1}{2}, and
    \end{equation}
  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
  \end{itemize}
  Then
  \begin{equation}
    \label{eqn:mnoisesuff}
    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
  \end{equation}
\end{proposition}

\begin{proof}
  From equation \ref{eqn:normsol},
  \[
    \aw[m;d](t) = 
    \left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2)\right)^{-1}\frac{1}{4 \pi r}d(t+mr)
  \]
  \[
    =4\pi r(1 + (4\pi r \alpha t)^2)^{-1} (\frac{1}{4\pi r}
    w_*(t+(m-m_*)r) + n(t+mr) 
  \]
  so
  \begin{equation}
    \label{eqn:wres}
    \aw[m;d](t) - w_*(t) = (4\pi r \alpha t)^2 (1+ (4\pi r \alpha t)^2)^{-1}w_*(t+(m-m_*)r)
    +w_*(t+(m-m_*)r)-w_*(t) + 4 \pi r (1 + (4\pi r \alpha t)^2)^{-1}n(t+mr)
  \end{equation}
\end{proof}
%  \begin{itemize}
%    \item[1. ] the bound \ref{eqn:mnoisebnd} holds only if 
%      noise level $\eta=\|n\|/\|d_*\|$ satisfies
%      \begin{equation}
%        \label{eqn:mnoisecond}
%        \eta < \frac{1+\sqrt{5}}{2} < 1;
%      \end{equation}
%    \item[2. ] if condition \ref{eqn:mnoisecond} holds, then
%      $|m-m_*|<\lambda/r$ provided that
%      \begin{equation}
%        \label{eqn:mnoisesuff}
%        \frac{\lambda}{\mu} \ge 1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}. 
%      \end{equation}
%      \begin{equation}
%        \label{eqn:mnoisesuff}
%        |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}. 
%      \end{equation}
%    \end{itemize}
%\end{corollary}

\begin{theorem}
  \label{thm:mnoiseres}
  Asumme that
  \begin{itemize}
  \item[1. ] $\alpha, \mu> 0$,
  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
  \item[3. ] $d_* = F[m_*]w_*$,
  \item[4. ] $n \in D$ and $d = d_* + n$.
  \end{itemize}
  Set $\eta = \|n\|/\|d_*\|$, and assume further that
  \begin{itemize}
  \item[5. ]
    \begin{equation}
      \label{eqn:mnoisecond}
      \eta < \frac{\sqrt{5}-1}{2}, and
    \end{equation}
  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
  \end{itemize}
  Then
  \begin{equation}
    \label{eqn:mnoisesuff}
    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
  \end{equation}
\end{theorem}

@@

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

\append{Numerics and penalty selection via the Discrepancy Principle}
This appendix describes a method for controlling the penalty
weight $\alpha$ based on the Discrepancy Principle
\cite[]{EnglHankeNeubauer,FuSymes2017discrepancy}.

@@

\noindent {\bf Remark:} Given a solution $m_{\infty}, \alpha_{\infty}$, the pair
$(m_{\infty},w_{\alpha_{\infty}}[m_{\infty};d])$ is a local minimizer
of $J_{\alpha_{\infty}}[\cdot,\cdot;d]$ (Theorem \ref{thm:norminv}).
%.
%so is also a minimizer of $g[w]$ over the set
%\[
%  \{(m,w): e[m,w;d] \ge
%  e[m_{\infty},w_{\alpha_{\infty}}[m_{\infty};d]]\}
%\]
%or equivalently a maximizer of $e[m,w,d]$ over the set
%\[
%  \{(m,w): m \in M, g[w] \le g[w_{\alpha_{\infty}}[m_{\infty};d]]\}
%\]
%according to standard Lagrange multiplier theory. That is,

@@

Given an
objective function $\Phi: M \rightarrow \bR$, a local optimization algorithm generates 
generates a map $G[\Phi,...]: M \rightarrow M$, mapping a current estimate
of $m$ to an updated estimate. The update rule may depend only on the
current estimate, as is the case for steepest descent, the
Gauss-Newton method, or Newton's method, or may also depend on
information generated during earlier updates, as for secant-type
methods such as BFGS. The notation is intended to allow for this
latter possibility. Under ``standard conditions'' on $\Phi$ and the initial estimate for
$m$, and equipped with so-called globalization safeguards to ensure
satisfaction of sufficient decrease conditions
(see \cite{NocedalWright}), the iteration produces a sequence in $M$
converging to a local minimizer of $\Phi$. An example of such a local
optimization algorithm is steepest descent with backtracking line
search.

@@

In outline, the algorithm is as follows:
\begin{algorithm}[H]
\caption{Scheme for updating $m, \alpha$}
\begin{algorithmic}[1]
  \State Choose $m\in M$
  \Repeat
  \State $\alpha \gets H[m,e_-,e_+]$
  \Repeat
  \State $m \gets G[\alpha,e_-,e_+,...](m)$
  \Until{$e[m,\aw[m,d];d] \notin [e_-,e_+]$ or $\|\nabla \tJa[m;d]\|$
    sufficiently small, or...}
  \Until{$e[m,\aw[m,d];d] \in [e_-,e_+]$} 
\end{algorithmic}
\end{algorithm}
Note that after step 3, the error bounds are satisfied, that is,
$e[m,\aw[m;d];d] \in [e_-,e_+]$, but after step 4, that is likely not
to be the case: the alteration of $m$ is likely to reduce $e$, as it is a
summand in the definition of $\tJa$. If the $e$ is reduced below
$e_-$, or if an approximate local minimizer is detected, then the condition in step 6 is satisfied, so control returns
to step 7. If it is satisfied, the algorithm terminates, elsewhich
control loops back to step 2, $\alpha$ is updated, and the inner $m$
update loop is entered again. Termination requires that the bounds on
$e$ are satisfied (step 7), hence that the $m$ update loop terminates
by finding a local min which satisfies these bounds.

Accordingly, regard $m$ as fixed and suppress it from the notation for the remainder of ;dthis section, and introduce the abbreviations
\begin{eqnarray}
\label{eqn:erralph}a
e(\alpha) &=& e[m,\aw[m;d],d]\\
\label{eqn:penalph}
p(\alpha) &=& p[m,\aw[m;d],d]
\end{eqnarray}
Since the normal matrix $F^T F + \alpha^2 A^T A$ has been assumed
invertible (condition \ref{eqn:nopco}), the solution of $\aw$ of the
normal equation \ref{eqn:condition} is smooth in $\alpha^2$.
Differentiating equation \ref{eqn:condition} with respect to  $\alpha^2$ leads to the relation
\begin{equation}
(F^T F + \alpha^2 A^T A ) \frac{dw}{d\alpha^2} = -A^T A w
\label{eqn:dnorm}
\end{equation}
whence
\begin{align}
\frac{de}{d\alpha^2} 
&=\left\langle\frac{dw}{d\alpha^2},F^T(Fw-d) \right\rangle \nonumber \\
&=-\alpha^2\left\langle\frac{dw}{d\alpha^2},A^TAw\right\rangle \nonumber \\ 
&=\alpha^2 \langle A^TAw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw\rangle \nonumber \\
&\ge 0
\label{eqn:de}
\end{align}
Note that the inequality in equation \ref{eqn:de} is {\em strict}  if $p > 0$ hence $A^TAw \ne 0$, since the normal operator is assumed to be positive definite.
The derivative of $p(\alpha^2)$ with respect to $\alpha^2$ is
\begin{align}
\frac{dp}{d\alpha^2} &=  -\langle A^T Aw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw \rangle \nonumber \\
&\leq 0
\label{eqn:dp}
\end{align}
similarly a strict inequality if $p > 0$.

>>>>

\title{Solution of an Acoustic Transmission Inverse Problem by Extended Inversion}
\author{William W. Symes\footnotemark[1], Huiyi Chen\footnotemark[2],
  and Susan E. Minkoff\footnotemark[2]}
\address{\footnotemark[1]PO Box 43, Orcas, WA 98280 USA\\
  \footnotemark[2]Department of Mathematical Sciences, FO 35, 
  University of Texas at Dallas, Richardson, TX 75080 USA}

@@

\section{Introduction}
Model-based parameter estimation by least squares data fitting is a
widely used and successful technique for data analysis in science and
engineering \cite{Bertero:98,Vogel:02}, particularly in the
geosciences \cite{Parker:94,Tarantola:05}. In its application to
seismology, least-squares data fitting has come to be known as Full
Waveform Inversion (FWI), and is now well-established as a useful tool
for probing the earth's subsurface
\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. However, in its
application to seismology, FWI encounters a serious
practical obstacle, so-called ``cycle-skipping''. This term signifies
the tendency of the mean-square error function to exhibit many
stationary points far from the global minimizer (we
show an explicit example of this phenomenon later in this paper). Because of the
typical dimensions of field FWI problems, only local gradient-based
minimization algorithms are computationally practical. These find
stationary points (or local minima), so may stagnate at suboptimal and geologically
uninformative earth models. Local descent methods avoid suboptimal
stagnation only if initial models are already quite close to optimal,
in the sense of predicting the arrival times of seismic events to
within a small multiple of the dominant data wavelength
\cite[]{GauTarVir:86,VirieuxOperto:09,Plessix:10}.

@@

\begin{corollary}
  \label{thm:epjnonoise}
  Assume the hypotheses of Proposition \ref{thm:norminvexp}, item
  3. Then noting that $\|d\| = \|w_*\|/(4 \pi r)$
\begin{equation}
  \label{eqn:residnorm}
  e[m,\aw[m,d];d] 
= \frac{\alpha^4}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
\end{equation}
\begin{equation}
  \label{eqn:anninorm}
  p[m,\aw[m,d];d] = \frac{(4\pi r)^2}{2\|w_*\|^2}\int \,dt\,  
  \frac{a(t-(m-m_*)r)^2}{(1+ (4\pi r)^2\alpha^2
    a(t-(m-m_*)r)^2)^{2}}w_*(t)^2.
\end{equation}
so 
\begin{equation}
\label{eqn:expjnonoise}
\tJa[m;d] = \frac{(4\pi r \alpha)^2}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
  a(t-(m-m_*)r)^2)^{-1}w_*(t)^2. 
\end{equation}
Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
\begin{equation}
  \label{eqn:dexpjnonoise}
  \frac{d}{dm}\tJa[m;d] = -\frac{r (4\pi r \alpha)^2}{\|w_*\|^2} \int \,dt \, 
  \frac{\left(a\frac{da}{dt}\right)(t-(m-m_*)r)}{(1+(4\pi r)^2 \alpha^2 
  a(t-(m-m_*)r)^2)^{2}}w_*(t)^2. 
\end{equation}
\end{corollary}

@@

\begin{proof}
  From the assumption $w_* \in W_{\mu}$ and Corollary
  \ref{thm:rampreallygood}, $|(m_{\infty}-m_*)r|\le \mu$. From the
  identity \ref{eqn:solnonoise},
  $\mbox{supp }\aw[m_{\infty};d] \subset
  [(m_{\infty}-m_*)r-\mu,(m_{\infty}-m_*)r+\mu] \subset
  [-2\mu,2\mu]$. Because of the support limitation, $a(t)=|t|$ in the
  interval of integration appearing in \ref{eqn:residnorm}, so
\[
  e[m_{\infty},\aw[m_{\infty},d];d] 
= 8 \pi^2 r^2 \alpha^4\int^{\mu}_{-\mu}\,dt\,\frac{|t-(m_{\infty}-m_*)r|^4}{(1+(4\pi r)^2 \alpha^2 
|t-(m_{\infty}-m^*)r|^2)^{2}}w_*(t)^2.
\]
and therefore
\begin{equation}
  \label{eqn:estresidnorm}
e[m_{\infty},\aw[m_{\infty};d];d] \le 8 \pi^2 r^2 (2\mu\alpha)^4 \|w_*\|^2 =
\frac{1}{2}(8 \pi r \mu \alpha)^4 \|d\|^2
\end{equation}
\end{proof}

@@

\begin{theorem}
  \label{thm:mnoiseres}
  Asumme that
  \begin{itemize}
  \item[1. ] $\alpha, \mu> 0$,
  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
  \item[3. ] $d_* = F[m_*]w_*$,
  \item[4. ] $n \in D$ and $d = d_* + n$.
  \end{itemize}
  Set $\eta = \|n\|/\|d_*\|$. Assume that $\eta$ satisfies inequality \begin{equation}
  \label{eqn:mnoisecond}    
  \eta < \frac{\sqrt{5}-1}{2},
  \end{equation}
  and that $m$ is a stationary point of $\tJa[\cdot;d]$.
  Then
  \begin{equation}
    \label{eqn:mnoisesuff}
    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
  \end{equation}
\end{theorem}

@@

\bibliographystyle{seg}
\bibliography{masterref}

\append{Penalty weight selection via the Discrepancy Principle}
This appendix describes a method for controlling the penalty
weight $\alpha$ based on a version the Discrepancy Principle
\cite[]{EnglHankeNeubauer,FuSymes2017discrepancy}.

@@

\noindent {\bf Remark:} Given a solution $m_{\infty}, \alpha_{\infty}$, the pair
$(m_{\infty},w_{\alpha_{\infty}}[m_{\infty};d])$ is a local minimizer
of $J_{\alpha_{\infty}}[\cdot,\cdot;d]$ (Theorem \ref{thm:norminv}).

@@

Given an
objective function $\Phi: M \rightarrow \bR$, a local optimization algorithm 
generates a map $G[\Phi,...]: M \rightarrow M$, mapping a current estimate
of $m$ to an updated estimate. The update rule may depend only on the
current estimate, as is the case for steepest descent, the
Gauss-Newton method, or Newton's method, or may also depend on
information generated during earlier updates, as for secant-type
methods such as BFGS. The notation is intended to allow for this
latter possibility. Under ``standard conditions'' on $\Phi$ and the initial estimate for
$m$, and equipped with so-called globalization safeguards to ensure
satisfaction of sufficient decrease conditions
(see \cite{NocedalWright}), the iteration produces a sequence in $M$
converging to a local minimizer of $\Phi$. An example of such a local
optimization algorithm is steepest descent with backtracking line
search.

@@

In outline, the algorithm is as follows:
\begin{algorithm}[H]
\caption{Scheme for updating $m, \alpha$}
\begin{algorithmic}[1]
  \State Choose $m\in M$
  \Repeat
  \State $\alpha \gets H[m,e_-,e_+]$
  \Repeat
  \State $m \gets G[\alpha,e_-,e_+,...](m)$
  \Until{$e[m,\aw[m,d];d] \notin [e_-,e_+]$ or $\|\nabla \tJa[m;d]\|$
    sufficiently small, or...}
  \Until{$e[m,\aw[m,d];d] \in [e_-,e_+]$} 
\end{algorithmic}
\end{algorithm}
Note that after step 3, the error bounds are satisfied, that is,
$e[m,\aw[m;d];d] \in [e_-,e_+]$, but after step 4, that is likely not
to be the case: the alteration of $m$ is likely to reduce $e$, as it is a
summand in the definition of $\tJa$. If the $e$ is reduced below
$e_-$, or if an approximate local minimizer is detected, then the condition in step 6 is satisfied, so control returns
to step 7. If it is satisfied, the algorithm terminates, else
control loops back to step 2, $\alpha$ is updated, and the inner $m$
update loop is entered again. Termination requires that the bounds on
$e$ are satisfied (step 7), hence that the $m$ update loop terminates
by finding a local min which satisfies these bounds.

Accordingly, regard $m$ as fixed and suppress it from the notation for the remainder of this section, and introduce the abbreviations
\begin{eqnarray}
\label{eqn:erralph}a
e(\alpha) &=& e[m,\aw[m;d],d]\\
\label{eqn:penalph}
p(\alpha) &=& p[m,\aw[m;d],d]
\end{eqnarray}
Assume that the normal operator $F^T F + \alpha^2 A^T A$ is boundedly invertible, as for example in Theorem \ref{thm:norminv}. Then the solution of $\aw$ of the
normal equation \ref{eqn:norm}
is smooth in $\alpha^2$.
Differentiating equation the normal equation  
with respect to  $\alpha^2$ leads to the relation
\begin{equation}
(F^T F + \alpha^2 A^T A ) \frac{dw}{d\alpha^2} = -A^T A w
\label{eqn:dnorm}
\end{equation}
whence
\begin{align}
\frac{de}{d\alpha^2} 
&=\left\langle\frac{dw}{d\alpha^2},F^T(Fw-d) \right\rangle \nonumber \\
&=-\alpha^2\left\langle\frac{dw}{d\alpha^2},A^TAw\right\rangle \nonumber \\ 
&=\alpha^2 \langle A^TAw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw\rangle \nonumber \\
&\ge 0
\label{eqn:de}
\end{align}
Note that the inequality in equation \ref{eqn:de} is {\em strict}  if $p > 0$ hence $A^TAw \ne 0$, since the normal operator is assumed to be positive definite.
The derivative of $p(\alpha^2)$ with respect to $\alpha^2$ is
\begin{align}
\frac{dp}{d\alpha^2} &=  -\langle A^T Aw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw \rangle \nonumber \\
&\leq 0
\label{eqn:dp}
\end{align}
similarly a strict inequality if $p > 0$.
